---
title: "Cyclist Bike Trip Analysis"
author: "Ajay Chaudhary"
date: "02/01/2022"
output: pdf_document
---

# Cyclist_Full_Year_Analysis

This analysis is based on cyclist case study.The purpose of this script is to consolidate downloaded Cyclist data into a single data frame and then conduct simple analysis to help answer the key question: “In what ways do annual members and casual riders use Cyclist bikes differently?

## Install Required packages

Install and load packages like tidyverse,lubridate and ggplot2

```{r loading packages,warning=FALSE, echo=TRUE, results='hide'}
library(tidyverse)
library(lubridate)
library(ggplot2)
```

## STEP 1: Upload data


Upload cyclist csv files:

```{r Upload CSV files,echo=TRUE,message=FALSE,results='hide'}
q2_2019 <- read_csv("D:/Coursera/Google_Analyze_data/Capstone_Project/Cyclist_Bike_Sharing/Cyclist_bike_trips-csv_files/Cyclist_Trips_2019_Q2.csv")
q3_2019 <- read_csv("D:/Coursera/Google_Analyze_data/Capstone_Project/Cyclist_Bike_Sharing/Cyclist_bike_trips-csv_files/Cyclist_Trips_2019_Q3.csv")
q4_2019 <- read_csv("D:/Coursera/Google_Analyze_data/Capstone_Project/Cyclist_Bike_Sharing/Cyclist_bike_trips-csv_files/Cyclist_Trips_2019_Q4.csv")
q1_2020 <- read_csv("D:/Coursera/Google_Analyze_data/Capstone_Project/Cyclist_Bike_Sharing/Cyclist_bike_trips-csv_files/Cyclist_Trips_2020_Q1.csv")
```


## STEP 2: WRANGLE DATA AND COMBINE INTO A SINGLE FILE


### Compare column names of each file

While the names don't have to be in the same order, they DO need to match perfectly before we can use a command to join them into one file

```{r column q2_2019, echo=TRUE}
colnames(q2_2019)
```
```{r column q3_2019}
colnames(q3_2019)
```
```{r column q4_2019}
colnames(q4_2019)
```
```{r column q4_2020}
colnames(q4_2019)
```


### Rename column

Rename columns  to make them consisent with q1_2020

```{r rename column,echo=TRUE,results='hide'}
(q4_2019 <- rename(q4_2019
                   ,ride_id = trip_id
                   ,rideable_type = bikeid 
                   ,started_at = start_time  
                   ,ended_at = end_time  
                   ,start_station_name = from_station_name 
                   ,start_station_id = from_station_id 
                   ,end_station_name = to_station_name 
                   ,end_station_id = to_station_id 
                   ,member_casual = usertype))
(q3_2019 <- rename(q3_2019
                   ,ride_id = trip_id
                   ,rideable_type = bikeid 
                   ,started_at = start_time  
                   ,ended_at = end_time  
                   ,start_station_name = from_station_name 
                   ,start_station_id = from_station_id 
                   ,end_station_name = to_station_name 
                   ,end_station_id = to_station_id 
                   ,member_casual = usertype))
(q2_2019 <- rename(q2_2019
                   ,ride_id = "01 - Rental Details Rental ID"
                   ,rideable_type = "01 - Rental Details Bike ID" 
                   ,started_at = "01 - Rental Details Local Start Time"  
                   ,ended_at = "01 - Rental Details Local End Time"  
                   ,start_station_name = "03 - Rental Start Station Name" 
                   ,start_station_id = "03 - Rental Start Station ID"
                   ,end_station_name = "02 - Rental End Station Name" 
                   ,end_station_id = "02 - Rental End Station ID"
                   ,member_casual = "User Type"))

```
 
 
### Inspect Dataframe


Inspect the dataframes and look for incongruencies

```{r Inspect Dataframe}
str(q1_2020)
```
```{r}
str(q4_2019)
```
```{r}
str(q3_2019)
```
```{r}
str(q2_2019)
```



### Convert Column format


Convert ride_id and rideable_type to character so that they can stack correctly

```{r Convert column format}
q4_2019 <- mutate(q4_2019, ride_id = as.character(ride_id), rideable_type = as.character(rideable_type))
q3_2019 <- mutate(q3_2019, ride_id = as.character(ride_id), rideable_type = as.character(rideable_type))
q2_2019 <- mutate(q2_2019, ride_id = as.character(ride_id), rideable_type = as.character(rideable_type))
```


### Join Dataframes


Join all the quarter data into one big data frame

```{r Join Dataframes}
all_trips <- bind_rows(q2_2019,q3_2019,q4_2019,q1_2020)
```


### Remove Columns


Remove lat, long, birthyear, and gender fields as this data was dropped beginning in 2020

```{r Remove Columns}
all_trips <- all_trips %>%
  select(-c(start_lat,start_lng,end_lat,end_lng,birthyear,gender,"01 - Rental Details Duration In Seconds Uncapped", "05 - Member Details Member Birthday Year", "Member Gender", "tripduration"))
```




## STEP 3: CLEAN UP AND ADD DATA TO PREPARE FOR ANALYSIS



### Inspect new table


Inspect the new table that has been created

list of column names

```{r column names}
colnames(all_trips)
```
Number of rows in data frame
```{r number of rows}
nrow(all_trips)
```
Dimension of the data frame
```{r Dimension}
dim(all_trips)
```
First six rows of data frame
```{r top rows}
head(all_trips)
```
List of columns and data types
```{r data types}
str(all_trips)
```
Statistical summary of data frame
```{r Summary}
summary(all_trips)
```

Number of observation fall under each usertype
```{r number of usertype}
table(all_trips$member_casual)
```



### Cleaning data


 In the "member_casual" column, replace "Subscriber" with "member" and "Customer" with "casual"
Before 2020, Cyclist used different labels for these two types of riders ... we will want to make our dataframe consistent with their current nomenclature

#### Reassign to the desired values (we will go with the current 2020 labels)


```{r reassign values}
all_trips <- all_trips %>%
  mutate(member_casual = recode(member_casual, "Subscriber" = "member", "Customer" = "casual"))
```


#### Adding columns


Add columns that list the date, month, day, and year of each ride

```{r Adding columns}
all_trips$date <- as.Date(all_trips$started_at)
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
```

Adding ride length column

```{r ride length}
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)
```

Inspect the structure of the column

```{r structure of the column}
str(all_trips)
```


#### Convert data format


Convert "ride_length" from Factor to numeric so we can run calculations on the data

```{r Convert data format}
is.factor(all_trips$ride_length)
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)
```


### Removing data

The dataframe includes a few hundred entries when bikes were taken out of docks and checked for quality by cyclist or ride_length was negative
We will create a new version of the dataframe (v2) since data is being removed

```{r Removing data}
all_trips_v2 <- all_trips[!(all_trips$start_station_name == "HQ QR"|all_trips$ride_length<0),]
```


## STEP 4: CONDUCT DESCRIPTIVE ANALYSIS


### Descriptive analysis on ride_length (all figures in seconds)

Summary of ride length

```{r Summary of ride length}
summary(all_trips_v2$ride_length)
```

### Compare members and casual riders

comparing mean, median, max, min in ride length between members and casual riders 

```{r Compare}
aggregate(all_trips_v2$ride_length~all_trips_v2$member_casual, FUN = mean)
aggregate(all_trips_v2$ride_length~all_trips_v2$member_casual, FUN = median)
aggregate(all_trips_v2$ride_length~all_trips_v2$member_casual, FUN = max)
aggregate(all_trips_v2$ride_length~all_trips_v2$member_casual, FUN = min)
```


See the average ride time by each day for members vs casual user

```{r average ride time by each day}
aggregate(all_trips_v2$ride_length~all_trips_v2$member_casual+all_trips_v2$day_of_week, FUN = mean)
```

Days of week are not in order.Let's fix that

```{r ordering days of week}
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels = c("Sunday","Monday","Tuesday","Wednesday","Thursday","Friday","Saturday"))
```

Now, let's run the average ride time by each day for members vs casual users

```{r}
aggregate(all_trips_v2$ride_length~all_trips_v2$member_casual+all_trips_v2$day_of_week, FUN = mean)
```


### Analyze ridership data by type and weekday

```{r Analyze data}
all_trips_v2 %>%
  mutate(weekday = wday(started_at, label = TRUE))%>%
  group_by(member_casual,weekday)%>%
  summarise(number_of_rides = n(),average_duration = mean(ride_length))%>%
  arrange(member_casual,weekday)

```



### Visualization


**Visualize the number of rides by the rider type**

```{r visualize the number of rides}
all_trips_v2 %>%
  mutate(weekday = wday(started_at,label = TRUE)) %>%
  group_by(member_casual,weekday) %>%
  summarise(number_of_rides = n(),average_duration = mean(ride_length)) %>%
  arrange(member_casual,weekday) %>%
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual))+
  geom_col(position = "dodge")
```



**Visualization for average duration**

```{r visualization for average duration}
all_trips_v2 %>%
   mutate(weekday = wday(started_at, label = TRUE))%>%
  group_by(member_casual,weekday)%>%
  summarise(number_of_rides = n(),average_duration = mean(ride_length))%>%
  arrange(member_casual,weekday) %>%
   ggplot(aes(x = weekday, y = average_duration, fill = member_casual))+
  geom_col(position = "dodge")
```

